---
title: "Results"
author: "Søren Wengel Mogensen"
date: "16 February 2016"
output: html_document
---


```{r}
library(xtable)
library(ggplot2)
library(reshape2)
load("../Main/mainResults1020")
```

### CV folds, ordered by iAUC

```{r cvfolds, echo = FALSE}
for (i in 1:length(res_list)) {
  colnames(res_list[[i]]) <- substr(colnames(res_list[[i]]), 1, 6)
  print(paste("CV fold", i))
  print(res_list[[i]], digits = 3)
}
```

### Average Performance (over CV folds)

```{r average1, echo = FALSE}
# re-order alphabetically to take averages
for (i in 1:length(res_list)) {
  res_list[[i]] <- (res_list[[i]])[order(rownames(res_list[[i]])), ]
}

addFunc <- function(x, ...) {
  ll <- list(...)
  for (i in 1:length(ll)) x <- x + ll[[i]]
  x
  }

ave_res <- do.call(addFunc, res_list)/length(res_list)
```


```{r average2, echo = FALSE}
spli <- unlist(strsplit(rownames(ave_res), split = "M"))
meth <- spli[seq_along(spli) %% 2 == 1]
impmeth <- paste("M", spli[seq_along(spli) %% 2 == 0], sep = "")

ave_df <- data.frame(Method = meth, Imputation = impmeth, ave_res)
```

```{r average3, results = "asis", echo = FALSE}
print(xtable(ave_df, digits = 3), type = "html", include.rownames = FALSE)
```


### Subtables for each criterion

```{r iAUCtable1, echo = FALSE}
# function for making a "two-sided" results table with only one criterion

subTable <- function(crit, dd = ave_df){
  
ss <- ave_df[ ,c("Method", "Imputation", crit)]
ssWide <- reshape(ss, v.names = crit, idvar = "Imputation",
                    timevar = "Method", direction = "wide")
rownames(ssWide) <- NULL
print(xtable(ssWide, digits = 3), type = "html", include.rownames = FALSE)
}

```

```{r iAUCtable2, results = "asis", echo = FALSE}
for (i in seq(ncol(ave_df) - 2)) { 
  print(names(ave_df)[i+2])
  subTable(crit = names(ave_df)[i+2])
}
```

### Average plot

Plot of the average results (over the CV folds) for each criterion, method and imputation method.

```{r resPlot1, echo = FALSE}
names(ave_df) <- substr(names(ave_df), 1, 6)
dd <- melt(ave_df, id.vars = c("Method", "Imputa"))
ggplot(dd, aes(x = variable, y = value, fill = Method)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  facet_grid( ~ Imputa) + coord_cartesian(ylim=c(.5, 0.8)) + 
  xlab(NULL) +
  theme(legend.position = "bottom")
```

Same plot as above, imputation and method have just switched roles to facilitate comparison between imputation methods.

```{r resPlot2, echo = FALSE}
ggplot(dd, aes(x = variable, y = value, fill = Imputa)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  facet_grid( ~ Method) + coord_cartesian(ylim=c(.5, 0.8)) + 
  xlab(NULL) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.3))
```

### iAUC for each fold

Plot of iAUC for each fold, method and imputation.

```{r foldPlot, echo = FALSE}
dd <- ave_df[ ,c(1,2)]
if (!identical(rownames(dd), rownames(res_list[[1]]))) 
  warning("Rows do not match, review code.")
for (i in 1:length(res_list)){
  dd[ ,i+2] <- res_list[[i]][ ,5]
}

names(dd) <- c("Method", "Imputation", paste("Fold", seq_along(res_list)))

dd <- melt(dd, id.vars = c("Method", "Imputation"))
ggplot(dd, aes(x = variable, y = value, fill = Method)) + 
  geom_bar(stat = "identity", position = position_dodge()) +
  facet_grid( ~ Imputation) + coord_cartesian(ylim=c(.5, 0.8)) +
  ylab("iAUC") + xlab(NULL) +
  theme(legend.position = "bottom")
```


### iAUC plot

```{r iAUCplot}
## levOrd <- names(sort(tapply(ave_df$iAUC, ave_df$Method, mean), decreasing = TRUE))
tmp <- ave_df[ave_df$Imputa == "MCAR", ]
tmp <- subset(tmp, Method != "gb")
ord <- order(tmp[, "iAUC"], decreasing = TRUE)
niceLabels <- c("Cox*", "Forest", "Forest*", "Gam*", "Boosting", 
                "Boosting*", "Debiased lasso", "Lasso")[ord]
levOrd <- tmp[ord, "Method"]
levOrd = levOrd[levOrd != "gb"]
dd2 <- dd
dd2 <- subset(dd2, Method != "gb")
dd2$Method <- factor(dd2$Method, levels = levOrd)
dd2$Imputa <- factor(dd2$Imputa, levels = c("MCAR", "MAR", "MARwR"))         
ave2 <- ave_df[, c(1, 2, 7)]
ave2 <- subset(ave2, Method != "gb")

postscript(file = "iAUC.eps", width = 10, height = 4,
           onefile = FALSE, horizontal = FALSE, paper = "special")
ggplot(dd2, aes(x = Method, y = value)) + 
  geom_point(color = "lightblue", size = 2) + 
  geom_point(aes(y = iAUC), data = ave2,
            color = "red", size = 3) + 
  facet_grid( ~ Imputa) + 
  coord_cartesian(ylim=c(.6, 0.75)) +
  ylab("Integrated AUC") + xlab("Method") +
  theme(legend.position = "top") + 
  theme(axis.title.y = element_text(margin = margin(0, 10, 0, 0))) +
  theme(axis.text.x  = element_text(angle = 45, vjust = 1, hjust = 1)) + 
  scale_x_discrete(labels = niceLabels)
dev.off()
```






